Visualizing Independence Using Extended Association Plots

### 2005-08-03 trunk/vcd/paper/JCGS.Rnw
essentially finished Section 2, started Section 3
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@426 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

remarks: Friendly 1994

A first extension by
\cite{vcd:Friendly:1994} uses a color coding of the tiles to visualize deviations
(residuals) from a given log-linear model fitted to the table, that is, from
the expected frequencies under independence. This approach does not only work in 2-way
tables but also in log-linear models fitted to multi-way tables. In this
extension, positive and negative signs of the residuals are coded by rectangles with
solid and dashed borders respectively. Furthermore, residuals exceeding an absolute
value of 2 are shaded light blue and red respectively, those that even
exceed an absolute value of 4 are shaded with full saturation. The heuristic
behind this shading is that
the Pearson residuals are approximately standard normal which implies
that the highlighted cells are those with
residuals \emph{individually} significant at approximately the 5\% and 0.01\%
level. But the main purpose of the shading is not to visualize significance but the \emph{pattern} of
deviation from independence \citep[p.~109]{vcd:Friendly:2000}. In particular,
this shading does not provide a visualization of the maximum test (\ref{eq:MaxAbs})
as in general it is unknown to which significance level $\alpha$ the
values $c_\alpha = 2$ and $4$ correspond for the table being visualized.

@@

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[htbp]
\begin{center}
<<shadings,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = c(2.5, 4, 2.5, 3), gp_args = list(eps = NULL))
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
mosaic(art, newpage = FALSE, gp = art_gp, margins = c(2.5, 3, 2.5, 4))
popViewport(2)
@
\caption{Mosaic plot for the Arthritis data with Friendly shading (left) and extended shading (right).}
\label{fig:shadings}
\end{center}
\end{figure}

@@

\cite{vcd:Friendly:1999} discusses grouping similar to coplots (conditioning
plots) \citep{vcd:Cleveland:1993} that lead to trellis graphics
(Becker, R.A., Cleveland, W.S. and Shyu, M. ``The Visual Design and
Control of Trellis Display'', Journal of Computational and
Graphical Statistics).

>>>>

Colors are commonly used to enhance mosaic and association plots. To integrate
a visualization of the Pearson residuals $\{r_{ij}\}$ into the mosaic display---which
in its `raw' version only visualizes the observed frequencies $\{n_{ij}\}$---\cite{vcd:Friendly:1994}
suggested a residual-based shading for the mosaic tiles that can also be applied
to the rectangles in association plots \citep{vcd:Meyer+Zeileis+Hornik:2003}. 
In this section, we first briefly review the \cite{vcd:Friendly:1994} shading, 
before we suggest different colors and a combination of visualization and significance
testing.

\subsection{Friendly Shading}

A first extension by
\cite{vcd:Friendly:1994} uses a color coding of the tiles to visualize deviations
(residuals) from a given log-linear model fitted to the table, that is, from
the expected frequencies under independence. This approach does not only work in 2-way
tables but also in log-linear models fitted to multi-way tables. In this
extension, positive and negative signs of the residuals are coded by rectangles with
solid and dashed borders respectively. Furthermore, residuals exceeding an absolute
value of 2 are shaded light blue and red respectively, those that even
exceed an absolute value of 4 are shaded with full saturation. The heuristic
behind this shading is that
the Pearson residuals are approximately standard normal which implies
that the highlighted cells are those with
residuals \emph{individually} significant at approximately the 5\% and 0.01\%
level. But the main purpose of the shading is not to visualize significance but the \emph{pattern} of
deviation from independence \citep[p.~109]{vcd:Friendly:2000}.

@@

In particular,
the Friendly shading does not provide a visualization of the maximum test (\ref{eq:MaxAbs})
as in general it is unknown to which significance level $\alpha$ the
values $c_\alpha = 2$ and $4$ correspond for the table being visualized.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[htbp]
\begin{center}
<<shadings,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = c(2.5, 4, 2.5, 3), gp_args = list(eps = NULL, lty = 1))
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
mosaic(art, newpage = FALSE, gp = art_gp, margins = c(2.5, 3, 2.5, 4))
popViewport(2)
@
\caption{Mosaic plot for the Arthritis data with Friendly shading (left) and extended shading (right).}
\label{fig:shadings}
\end{center}
\end{figure}

@@

\cite{vcd:Friendly:1999} discusses grouping similar to coplots (conditioning
plots) \citep{vcd:Cleveland:1993} that lead to trellis graphics
\citep{vcd:Becker+Cleveland+Shyu:1996}.
